---
title: "Final Project Report"
author:
  - name: "Leila Aburomia"
    corresponding: true
  - name: "Aldair Perez"
  - name: "Noah Goodhart"
format:
  html: 
    code-fold: true
execute: 
  echo: true
---

## El Niño-Southern Oscillation and Drought Linkages in Colorado Counties

# Abstract

The western United States has faced significant hydroclimatic variability in recent decades, with repeated drought events in the 1930s and 1950s. In this study, we use the Standardized Precipitation Index (SPI) and the Drought Severity and Coverage Index (DSCI) to examine linkages between drought severity in Colorado and the El Niño-Southern Oscillation (ENSO). The analysis will focus on five Colorado counties-Kit Carson, Costilla, Larimer, Mesa, and Park. We utilized R to preform correlation analyses and linear regression modeling to determine how ENSO severity correlates with precipitation and drought patterns. Our results show a weak to moderate negative correlation between ENSO severity and SPI in most counties, with Costilla being the exception. Drought Severity and Coverage Index exhibited little to no statically significant correlation across all counties, suggesting that ENSO may have little predictive power for drought severity during it's specific cycle. This finding apposes the common assumption that ENSO increases precipitation in Colorado during El Niño events. We observed a potential temporal lag between ENSO phases and present drought conditions, suggesting that insight may be improved by lagged time-series modeling. This examination highlights the complexity of climate-drought interactions in mountainous regions of the U.S. and demonstrates the importance of long-term data science in environmental research. Potential future work will explore the lagged effects of ENSO on drought conditions to improve understanding for future climate variability.

# Introduction

Over the past decade, the western United States has experienced hydroclimatic variability that has emerged as a dominant environmental issue of great concern (@Balling2007). Since 1996, a considerable portion of the western U.S. has seen drought conditions that are similar in scale to the historical droughts in the 1930s and 1950s (@Balling2007). Studies have also used the paleoclimatic record to reveal evidence of long term droughts over the past 500 to 2,000 years (@Woodhouse1998). Hydroclimatic variability studies range in scale from regional to watershed. This variability is often complex and may require a multitude of scales to fully understand and plan for future drought conditions. In the dry western U.S., Colorado receives less than 17 inches of average annual precipitation statewide. (@Mckee2000). In comparison to the rest of the United States, these rainfall amounts are incredibly sparse. Luckily, Colorado’s Rocky Mountains help alleviate potential water stress by collecting more than 25 inches of precipitation annually (@Mckee2000). By successfully adapting to this dry climate, Colorado supports a massive agricultural and recreational industry and multiple growing urban centers (@Mckee2000). As population and industry grow, the threat of drought and its social, economic, and environmental impacts increase as well.

Measuring and assessing drought accurately is essential for future planning. Historical drought data can give valuable insight into influences or indicators of drought conditions. Patterns of climatic variability that aligned with droughts in the past can help our society better prepare for the future. One of these patterns is the influential El Niño and La Niña. El Niño is caused by a warming of the tropical Pacific Ocean that occurs about every three to seven years and lasts for 12-18 months (@McPhaden2002). La Niña is akin to El Niño with stronger than normal trade winds and cooler tropical Pacific water temperatures (@McPhaden2002). In recent decades, the El Niño /Southern Oscillation (ENSO) has been more closely monitored, and the strength of each cycle has been documented. The effects of El Niño and La Niña on climate vary considerably across the globe, but in the western U.S., there is a pattern of impacts. During El Niño years, both extremely wet and dry conditions can be expected, while La Niña years are characterized by only dry conditions (@Canon2007). This research aims to analyze interactions between climatic variability indicating drought and the ENSO cycle. To analyze potential relationships we will look at correlations within 4 Colorado counties; Kit Carson, Costilla, Larimer, Mesa, and Park. To examine interactions between drought conditions and the El Niño /Southern Oscillation (ENSO), we will utilize a variety of climatic data.

To create a framework for drought, we will use the Standardized Precipitation Index (SPI) from the National Integrated Drought Information System to measure how much observed precipitation deviated from the climatological average. This condition can be measured on a variety of time scales, indicating that the SPI can be useful for both short-term and long-term applications (@NASA). The unit of SPI can be interpreted as “standard deviations”. We will also use the Drought Monitor of Colorado from the USDM (Drought Severity and Coverage Index) to assess the severity and extent of drought using their standardized classification system. This DSCI data uses six classifications to characterize drought severity: normal conditions, abnormally dry (D0), and four levels of drought (D1 to D4) representing moderate, severe, extreme, and exceptional drought conditions, respectively. In order to compare this data to ENSO cycles, we will examine El Niño and La Niña intensities for each year based on the Oceanic Niño Index (ONI). This index is the standard for NOAA used to classify El Niño and La Niña events in the eastern tropical Pacific (@Null). Events are defined as five consecutive 3-month periods at or above the + 0.5 degrees threshold for warm (El Niño ) events and at or below the - 0.5 degrees anomaly for cool (La Nina) events (@Null). 

To find correlations between our drought framework and ENSO cycles, we will use R software to conduct statistical tests. First, we will preprocess our data by aligning data sets by date. For SPI and DSCI we will standardize to 3 month rolling averages in order for our data to align with ENSO data. Three month rolling averages are beneficial in revealing more meaningful trends in drought conditions and precipitation anomalies. Our statistical methods will include a correlation analysis to test correlations between ENSO index (ONI) and drought variables. We will use the Pearson correlation coefficient to measure the linear relationship within the 4 Colorado counties. Then we will use ggplot2 to visualize the trends in a scatter plot and add a linear regression line to help visualize the relationship. If the correlation is significant, we will perform a correlation hypothesis test to provide us with the correlation coefficient and p-value. Next, we will conduct a time series analysis of cross-correlation to look for effects of ENSO on drought parameters. These findings will help us find a relationship between El Niño/La Niña and drought in Colorado. Considering we are living in a society becoming more conscious about climate change and its environmental impacts, such as more extreme drought events, this project will enforce our involvement in the understanding aspects of our global environment and comprehend to an extent on how global processes impact extreme weather. This project aims to examine the relationship between ENSO (El Niño -Southern Oscillation) using the Oceanic Niño Index and drought conditions in 5 Colorado counties (Kit Carson, Costilla, Larimer, Mesa, and Park) using SPI and DSCI data. Our hypothesis states that there is a statistically significant correlation between ENSO phases and drought severity in Colorado counties, with La Niña conditions associated with increased drought severity and El Niño conditions associated with more variable precipitation.

# Methods

This research utilizes a multi-step data science approach to examine the relationship between drought variability and ENSO phases across five Colorado counties. We integrated data sets from multiple sources, including ENSO data from Oceanic Niño Index (ONI) (@Null), precipitation data in the form of Standardized Precipitation Index (SPI) from drought.gov (@NOAA) and Drought Severity and Coverage Index (DSCI) from US drought monitor (@Simeral2025).

There are several methods we have chosen and implemented in our analysis. To begin, ENSO, SPI, and DSCI data were combined into one data set separated by county. each row is based off a three month rolling average of drought data points. The data ranged from the beginning of 2020 to the October 1st, 2024. Data preprocessing is as follows. Row #1 was changed to column names and converted the date from numeric to date from our dataset. Secondly, the data was cleaned and reshaped by choosing the most important columns (SPI, DSCI, ENSO, and Date), changing the names (ENSO severity to type, adding underscores to all column names), and converting severity and value from character to numeric format. The data was converted from wide format (many columns for different counties) to long format, creating two new columns, Variable (the original column names) and Value (the actual data values). The variable column was split into Metric_type and County.

R code was created and saved a plot to visualize the relationship between ENSO severity and SPI per county. From the long-form dataset, only kept rows where the metric type is Rolling SPI Average. The scatter plot showed the X-axis as ENSO severity and the y-axis as the rolling SPI average, with color representing different counties. Individual data points and a linear regression line were added to the plot. The axis tick labels were formatted for readability and 'theme_minimal()' was chosen for a simple plot theme. Lastly, the finished plot was added to the 'images' folder as a PNG file. This process was repeated to visualize the relationship between ENSO and DSCI.

A statistical test was run using Pearson to find a correlation between each county (Kit Carson, Costilla, Larimer, Mesa, Park) and ENSO severity (include Value) and run a linear regression model for each county. To calculate Pearson correlation for SPI, we filtered the data to use only SPI values and group the data by County. This was repeated for DSCI by filtering for DSCI instead of SPI. The linear regression model was run by again filtering for either SPI or DSCI first and grouping by County. The linear regression model was run using SPI value (or DSCI) as the response variable and ENSO severity as the predictor.

Time series modeling was implemented. A left join was created (date and rolling 3 month date), created a “center date” for each rolling period (where each acronym is assigned to a number that correlates to the month), and mutating the center month, year, and rolling date.

For time series visualizations, 'mutate()' was used to categorize each ENSO phase/type to either El Nino or La Nina. ENSO bands were created by selecting the rolling data and the categorization of ENSO, distinct, arrange by rolling date, added a column where a lag was run based on the ENSO phase, group by ENSO phase, and summarise (min and max are rolling date, where the group is “drop”. Visualizations were made for each relationship (SPI and DSCI for ENSO), where the X-axis is the date and y-axis is the rolling SPI/DSCI average, ENSO phase was implemented for each section of the time (El Nino, La Nina, neutral, NA), and the counties were implemented. The graphs were faceted that show 5 graphs, each representing each county.

Each method was implemented correctly based on dataset processing, using the most important sections of the data, generating graphs to find a relationship, running statistical tests to have a concise number for understanding the data and relationship, and then implementing time series models for both relationships.

# Results

```{r}
library(here)
library(readxl)

# import Dataset as Tibble
data1 <- read_excel(here("data", "Final_Data (1).xlsx"), col_names = FALSE)

# change row 1 into column names
colnames(data1) <- as.character(data1[1, ])
data1 <- data1[-1, ]

# Convert date (for reference column) from numeric to Date 
data1$`Date (for Reference)` <- as.Date(as.numeric(data1$`Date (for Reference)`), origin = "1899-12-30")
```

### Clean and Reshape Data

```{r}
library(dplyr)
library(tidyr)

# Select only the relevant columns (SPI/DSCI + ENSO + Date)
long_data <- data1 %>%
  select(`Date (for Reference)`, `ENSO Type`, `ENSO Severity`,
         starts_with("Rolling SPI Average"), starts_with("Rolling DSCI Average")) %>%
  pivot_longer(cols = starts_with("Rolling"),
               names_to = "Variable",
               values_to = "Value") %>%
  extract(Variable, into = c("Metric_Type", "County"), regex = "^(.*) ([^ ]+)$")

colnames(long_data) <- gsub(" ", "_", colnames(long_data))

# swap ENSO severity with ENSO type
long_data <- long_data %>%
  rename(`Temp_Column` = `ENSO_Severity`)

long_data <- long_data %>%
  rename(
    `ENSO_Severity` = `ENSO_Type`,
    `ENSO_Type` = `Temp_Column`
  )
# add underscores in all column names 
colnames(long_data) <- gsub(" ", "_", colnames(long_data))

# convert ENSO_severity and value to numeric
long_data$ENSO_Severity <- as.numeric(long_data$ENSO_Severity)
long_data$Value <- as.numeric(long_data$Value)
```

### Visualization of relationships between ENSO severity and SPI/DSCI per county

#### ENSO severity and Rolling SPI Average

```{r}
library(ggplot2)
library(scales)
library(dplyr)

ENSO_Rolling_SPI_plot <- ggplot(long_data %>% filter(Metric_Type == "Rolling SPI Average"),
       aes(x = `ENSO_Severity`, y = Value)) +
  geom_point(aes(color = County)) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Relationship between ENSO Severity and Rolling SPI Average",
       x = "ENSO Severity", y = "Rolling SPI Average") +
   scale_y_continuous(labels = number_format(accuracy = 0.01)) +  
  scale_x_continuous(labels = number_format(accuracy = 0.1)) +   
  theme_minimal() +
  facet_wrap(~ County)

ENSO_Rolling_SPI_plot


ggsave("images/ENSO_Rolling_SPI_plot.png", plot = ENSO_Rolling_SPI_plot, width = 6, height = 4, dpi = 300)
```

North (-0.37), East (-0.14), West (-0.24), and Central (-0.32) counties (Larimer, Kit Carson, Mesa, and Park, respectfully), show a negative weak to moderate correlation between ENSO severity and rolling SPI average. The Southern county, Costilla (0.08) exhibited a weak positive correlation between ENSO severity and rolling SPI average.

#### ENSO Severity and Rolling DSCI Average

```{r}
ENSO_Rolling_DSCI_plot <- ggplot(long_data %>% filter(Metric_Type == "Rolling DSCI Average"),
       aes(x = `ENSO_Severity`, y = Value)) +
  geom_point(aes(color = County)) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Relationship between ENSO Severity and Rolling DSCI Average",
       x = "ENSO Severity", y = "Rolling DSCI Average") +
   scale_y_continuous(labels = number_format(accuracy = 0.01)) +  
  scale_x_continuous(labels = number_format(accuracy = 0.1)) +   
  theme_minimal() +
  facet_wrap(~ County)


ENSO_Rolling_DSCI_plot

ggsave("images/ENSO_Rolling_DSCI_plot.png", plot = ENSO_Rolling_DSCI_plot, width = 6, height = 4, dpi = 300)
```

Mesa (0.07) and Costilla (0.02) county show a weak positive correlation between ENSO severity and rolling DSCI average. Larimer (-0.07) and Kit Carson (-0.1) county show a weak negative correlation, and Park county (-0.004) shows practically no correlation between ENSO severity and rolling DSCI average.

### Statistical Tests

```{r}
# Pearson correlation by county SPI
SPI_cor_results <- long_data %>%
  filter(Metric_Type == "Rolling SPI Average") %>%
  group_by(County) %>%
  summarise(correlation = cor(`ENSO_Severity`, Value, use = "complete.obs"),
            .groups = "drop")

print(SPI_cor_results)


# Linear Regression by county SPI
SPI_lm_results <- long_data %>%
  filter(Metric_Type == "Rolling SPI Average") %>%
  group_by(County) %>%
  group_map(~ summary(lm(Value ~ `ENSO_Severity`, data = .x)))

print(SPI_lm_results)
```

Kit Carson County (East) had a p-value of 0.0152 (significant), which suggests that ENSO influences the Standardized Precipitation Index (SPI) in this area, but with an R2 value of 0.02, there is weak explanatory power. Costilla County (South) had a p-value of 0.174 (not significant), which suggests that SPI is slightly affected by ENSO severity, but it’s not a statistically significant difference. The R2 value of 0.006 shows very weak explanatory power. Larimer County (North) had a p-value of 0.0000000003 (very significant), which suggests a strong negative association. It's R2 value of 0.14, shows moderate explanatory power of ENSO severity on SPI. Mesa County (West) had a p-value of 0.000025 (very significant), which suggests a strong negative association. With an R2 value of 0.058, there is low explanatory power, but meaningful.For Park County (Central), we got a p-value of 0.000000018 (very significant), which suggests a strong negative association. Again, higher ENSO severity leads to lower SPI. With an R2 value of 0.10, there is low to moderate explanatory power.

```{r}

# Pearson correlation by county DSCI
DSCI_cor_results <- long_data %>%
  filter(Metric_Type == "Rolling DSCI Average") %>%
  group_by(County) %>%
  summarise(correlation = cor(`ENSO_Severity`, Value, use = "complete.obs"),
            .groups = "drop")

print(DSCI_cor_results)

# Linear Regression by county DSCI
DSCI_lm_results <- long_data %>%
  filter(Metric_Type == "Rolling DSCI Average") %>%
  group_by(County) %>%
  group_map(~ summary(lm(Value ~ `ENSO_Severity`, data = .x)))

print(DSCI_lm_results)
```
Kit Carson County (East) had a p-value of 0.0814 (not significant) and an R2 value of 0.01, with only 1% of DSCI variability explained by ENSO. Costilla County (South) had a p-value of 0.769 (not significant). With an R2 value of 0, there is essentially no relationship. Larimer County (North) had a p-value of 0.231 (not significant) suggesting a non significant relationship. With an R2 value of 0.005, less than 1% of variability is explained. Mesa County (West) had a p-value of 0.207 (not significant). With an R2 value of 0.05, there is low explanatory power. Park County (Central) had a p-value of 0.932 (not significant). With an R2 value of 0, there is 0 explanatory power.

### Time Series Plots

```{r}
# more data cleaning

library(dplyr)

data1 <- data1 %>%
  rename(Date = `Date (for Reference)`)

long_data <- long_data %>%
  rename(Date = `Date_(for_Reference)`)

long_data <- long_data %>%
  left_join(data1 %>% select(Date, Rolling_3_Month_Date = `Rolling 3 Month Date`), by = "Date")
```

#### Create a "Center Date" for the Rolling 3-Month Period

```{r}
# Create a lookup table for center months of each rolling period
period_centers <- c(
  "JFM" = "02", "FMA" = "03", "MAM" = "04", "AMJ" = "05",
  "MJJ" = "06", "JJA" = "07", "JAS" = "08", "ASO" = "09",
  "SON" = "10", "OND" = "11", "NDJ" = "12", "DJF" = "01"
)

# Add a "centered" date column
long_data <- long_data %>%
  mutate(
    center_month = period_centers[`Rolling_3_Month_Date`],
    year = lubridate::year(Date),
    rolling_date = as.Date(paste(year, center_month, "15", sep = "-"))
  )
```

#### Time Series Visualizations for SPI and ENSO

```{r}
long_data <- long_data %>%
  mutate(
    ENSO_Phase = case_when(
      ENSO_Type %in% c("WE", "ME", "SE", "VSE") ~ "El Nino",
      ENSO_Type %in% c("WL", "ML", "SL") ~ "La Nina",
      ENSO_Type == "NONE" ~ "Neutral",
      TRUE ~ NA_character_ 
    )
  )

enso_bands <- long_data %>%
  select(rolling_date, ENSO_Phase) %>%
  distinct() %>%
  arrange(rolling_date) %>%
  mutate(group = cumsum(ENSO_Phase != lag(ENSO_Phase, default = first(ENSO_Phase)))) %>%
  group_by(group, ENSO_Phase) %>%
  summarise(start = min(rolling_date), end = max(rolling_date), .groups = "drop")

ggplot() +
  geom_rect(data = enso_bands,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf, fill = ENSO_Phase),
            alpha = 0.2) +
  # SPI lines
  geom_line(data = long_data %>% filter(Metric_Type == "Rolling SPI Average"),
            aes(x = rolling_date, y = Value, color = County)) +
  scale_fill_manual(values = c("El Nino" = "red", "La Nina" = "blue", "Neutral" = "gray")) +
  labs(title = "SPI with ENSO Phase Overlays",
       x = "Date", y = "Rolling SPI Average",
       fill = "ENSO Phase", color = "County") +
  theme_minimal()

# Faceted Trends 
Faceted_Time_Series_SPI <- ggplot() +
  # ENSO shaded bands across all facets
  geom_rect(data = enso_bands,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf, fill = ENSO_Phase),
            alpha = 0.2, inherit.aes = FALSE) +
  # SPI time series line
  geom_line(data = long_data %>% filter(Metric_Type == "Rolling SPI Average"),
            aes(x = rolling_date, y = Value, color = County)) +
  facet_wrap(~County, scales = "free_y") +
  scale_fill_manual(values = c("El Nino" = "red", "La Nina" = "blue", "Neutral" = "gray")) +
  labs(title = "SPI Over Time with ENSO Phase Overlays",
       x = "Date", y = "Rolling SPI Average",
       fill = "ENSO Phase", color = "County") +
  theme_minimal()

Faceted_Time_Series_SPI

ggsave("images/Faced_Time_Series_SPI.png", plot = Faceted_Time_Series_SPI, width = 10, height = 4, dpi = 300)

```

#### Time Series Visualizations for DSCI and ENSO

```{r}
Faceted_Time_Series_DSCI <- ggplot() +
  geom_rect(data = enso_bands,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf, fill = ENSO_Phase),
            alpha = 0.2, inherit.aes = FALSE) +
  geom_line(data = long_data %>% filter(Metric_Type == "Rolling DSCI Average"),
            aes(x = rolling_date, y = Value, color = County)) +
  facet_wrap(~County, scales = "free_y") +
  scale_fill_manual(values = c("El Nino" = "red", "La Nina" = "blue", "Neutral" = "gray")) +
  labs(title = "DSCI Over Time with ENSO Phase Overlays",
       x = "Date", y = "Rolling DSCI Average",
       fill = "ENSO Phase", color = "County") +
  theme_minimal()

ggsave("images/Faced_Time_Series_DSCI.png", plot = Faceted_Time_Series_DSCI, width = 10, height = 4, dpi = 300)

```

# Discussion

-   Looking at the visualization for the relationship between ENSO Severity and Rolling SPI Average, we can see that for our North, East, West, and Central counties (Larimer, Kit Carson, Mesa, and Park, respectfully), there is a negative weak to moderate correlation between ENSO severity and SPI. However our Southern County, Costilla, has a weak positive correlation between the two variables. When we run a linear regression on our five counties to identify if ENSO severity can predict SPI values, are results are as follows:

    1.  For Kit Carson County (East), we got a p-value of 0.0152 (significant), which suggests that ENSO drives drier conditions in this area, but with an R2 value of 0.02, there is weak explanatory power.

    2.  For Costilla County (South), we got a p-value of 0.174 (not significant), which suggests that SPI seems to increase with ENSO severity, but it’s not statistically significant. This could be due to noise or local variability. With an R2 value of 0.006, there is very weak explanatory power.

    3.  For Larimer County (North), we got a p-value of 0.0000000003 (very significant), which suggests a strong negative association. ENSO events have a marked drying effect in this county. With an R2 value of 0.14, there is moderate explanatory power.

    4.  For Mesa County (West), we got a p-value of 0.000025 (very significant), which suggests a strong negative association. Higher ENSO severity leads to lower SPI. With an R2 value of 0.058, there is low explanatory power, but meaningful.

    5.  For Park County (Central), we got a p-value of 0.000000018 (very significant), which suggests a strong negative association. Again, higher ENSO severity leads to lower SPI. With an R2 value of 0.10, there is low to moderate explanatory power.

-   We then ran the same statistical analyses (correlation test, and linear regression) to determine the relationship between ENSO severity and DSCI (drought) values. Before going through the results of our linear regression, it is important to look at our visualization of the data. We can see from the graph that Costilla and Mesa County have a slight positive correlation between ENSO severity and DSCI values, Larimer and Carson County have a slight negative correlation between the two variables, and Park county has practically no correlation. Here are the results of our linear regression:

    1.  For Kit Carson County (East), we got a p-value of 0.0814 ( borderline significant), which suggests that there is a possible weak drought & ENSO connection here. Worth noting but not strong. An R2 value of 0.01, only 1% of DSCI variability is explained by ENSO.
    2.  For Costilla County (South), we got a p-value of 0.769 (not significant at all), which suggests that ENSO doesn’t appear to affect DSCI in this county. With an R2 value of 0, there is essentially no relationship.
    3.  For Larimer County (North), we got a p-value of 0.231 (not significant), which suggests a very weak, nonsignificant relationship. ENSO events have little to no relationship to DSCI. With an R2 value of 0.005, this explains less than 1% of variability.
    4.  For Mesa County (West), we got a p-value of 0.207 (not significant), which again suggests a very weak nonsignificant relationship. Higher ENSO severity has barely any effect on DSCI values. With an R2 value of 0.05, there is low explanatory power.
    5.  For Park County (Central), we got a p-value of 0.932 (very far from significant), which suggests absolutely 0 relationship between ENSO and DSCI values. With an R2 value of 0, there is 0 explanatory power.

Our data implies that there is an overall negative association between ENSO severity and SPI values, and no significant relationship between ENSO severity and DSCI values. This goes against what we previously believed. We were under the assumption that a higher ENSO value (Stronger El Niño seasons), would result in higher precipitation values, leading to reduced drought conditions (@Bao2024). Our initial assumption is true for the southern part of Colorado (Costilla County), but after looking at the statistical analyses, we can see that this correlation between both ENSO & SPI, and ENSO & DSCI is very weak, and not significant.

There are a few limitations to our analysis, but one is most prominent. There is most likely a lag between ENSO severity seasons, and drought conditions being present. For example, if there is a strong La Niña event in January of a particular year, it would be around 3 months before drought is reported. There also might be a shorter lag between ENSO severity and precipitation measurements. Other limits may include: small sampling size. Including more counties might get us a better picture. To clean the data, we had to average 3 or 4 monthly drought values to line up with our ENSO severity values, and SPI values. There may be data hidden in a more complex analysis.

Our findings go against findings presented in an article published by NOAA. In the article, the author states that during La Niña years, the southwest tends to experience more dry conditions compared to El Niño years. However, looking at interactive graphs linked to the article, Colorado doesn’t seem to follow trends of drought seen in other southwestern states. This could explain why our data is pointing to a weak, negative correlation between ENSO severity, and drought and precipitation patterns.

Our findings connect directly to data science by modeling real world scenarios. We are using core data science tools, like R and tidyverse to analyze patterns across regions, interpret uncertainty through p-values and R² values, and communicate nuanced findings. This approach is central to environmental data science and climate analytics, supporting decisions in sustainability and resource management. Our work reflects key skills in data wrangling, statistical modeling, and scientific communication which are all essential to modern data science practice.

Looking forward, we are going to experiment with the lag function in R to possibly get a more accurate representation of the relationship between ENSO severity and DSCI values. We will have to be careful as to not manipulate the data, so it fits the narrative we want. We will have to research the exact lag between precipitation conditions and drought conditions, apply the lag function, and report our finding, whatever they happen to be.
